---
title: "topotools"
output: github_document
---

Coding tools for working with local copies of [ETOPO data](https://ngdc.noaa.gov/mgg/global/) and [GEBCO data](https://www.gebco.net).

### ETOPO Citation
```
Amante, C. and B.W. Eakins, 2009. ETOPO1 1 Arc-Minute Global Relief Model:
Procedures, Data Sources and Analysis. NOAA Technical Memorandum NESDIS NGDC-24.
National Geophysical Data Center, NOAA. doi:10.7289/V5C8276M [2018-11-01].
```

### GEBCO Citation

```
GEBCO Compilation Group (2020) GEBCO 2020 Grid (doi:10.5285/c6612cbe-50b3-0cff-e053-6c86abc09f8f).
```

## Requirements

+ [terra](https://CRAN.R-project.org/package=terra)

+ [ncdf4](https://CRAN.R-project.org/package=ncdf4)

## Installation

```
devtools::install_github("BigelowLab/topotools")
```

## Usage

```{r load_packages, message = FALSE}
library(terra)
library(topotools)
library(viridisLite)
```

### Paths

The root path to the topography data defaults to that which works for our setup. You can easily override this - see the `read_gebco` and `read_etopo` functions. 
``` {r setup}
library("topotools")
dir(topo_path(), full.names = TRUE)
```

### Read in a region

While it is possible to read in the entire dataset for each source, generally the practice is to read in a portion defined by a bounding box specified in `[west, east, south, north]` order.

```{r read}
bb <- c( -72,  -63,   39,   46)

etopo <- read_etopo(bb = bb)
gebco <- read_gebco(bb = bb)

etopo
gebco
```
Note that the GEBCO data provides approximately 4x the resolution of the ETOPO1 data.

For display purposes, it is helpful to clip each raster into a small range of values.

```{r plot}
par(mfrow = c(1,2))
plot(etopo, 
     col = viridisLite::viridis(50), 
     main = "ETOPO1", 
     range = c(-400, 800), 
     legend = FALSE)
plot(gebco, 
     col = viridisLite::viridis(50), 
     main = "GEBCO", 
     range = c(-400, 800), 
     legend = FALSE)
par(mfrow = c(1,1))
```
### Masking

The `mask_topo` function will work with either data set (well, any `terra::SpatRaster` or `raster::RasterLayer` object.)  Use that function in conjuction with `terra::mask` to produce a masked raster.

```{r mask}
par(mfrow = c(1,3))
masked_land <- terra::mask(etopo, topotools::mask_topo(etopo, where = "above"))
masked_sea <- terra::mask(etopo, topotools::mask_topo(etopo, where = "below"))
plot(etopo, 
     col = viridisLite::viridis(50), 
     main = "ETOPO1", 
     range = c(-400, 800), 
     legend = FALSE,
     axes = FALSE,
     mar = NA)
plot(masked_land, 
     col = viridisLite::viridis(50), 
     main = "Masked Land ETOPO1", 
     range = c(-400, 800), 
     legend = FALSE,
     axes = FALSE,
     mar = NA)
plot(masked_sea, 
     col = viridisLite::viridis(50), 
     main = "Masked Ocean ETOPO1", 
     range = c(-400, 800), 
     legend = FALSE,
     axes = FALSE,
     mar = NA)
par(mfrow = c(1,1))
```
